---
title: "routine-individual-diffs-analysis-lt"
format: html
editor: visual
---

## How does the formation of routine impact learning transfer?

We have formulated a measure of routine which we have calculated per person. We now want to understand how this impacted their capacity to transfer learning in the final stage.

```{r}
# first load some helpful libraries
library(tidyverse)
library(GGally)
library(ez)
library(wesanderson)
library(vioplot)
```

### Sorting data

First I will load and join the relevant data files. First the test data.

```{r}
exp_str = 'lt'

test_dat = read.csv(file=paste("../../doors-data/data-wrangled/exp", exp_str, "avg.csv", sep="_"))
test_dat = test_dat %>% filter(ses == 3) %>% select(sub, train_type, ses, transfer, accuracy)
test_dat = test_dat %>% group_by(sub, train_type) %>% summarise(acc_ratio = accuracy[transfer == 1] / accuracy[transfer == 2]) %>% ungroup()
head(test_dat)

```

Next the Maggi K4 data

```{r}

k4_dat = read.csv(file=paste("../../doors-data/data-wrangled/exp", exp_str, "maggi-k4.csv", sep="_")) %>% select(sid, ses, transfer, k4_onset) %>% filter(ses == 3) %>%
  group_by(sid, ses) %>% summarise(k4_ratio = k4_onset[transfer == 1] / k4_onset[transfer == 2])
names(k4_dat)[names(k4_dat) == "sid"] = "sub"
head(k4_dat)
```

And now the routine measure data

```{r}

r_dat = read.csv(file=paste("../../doors-data/data-wrangled/exp", exp_str, "rscore.csv", sep="_"))
head(r_dat)
```

Now I'll join the data all together

```{r}

all_dat <- inner_join(test_dat, k4_dat %>% select(sub, k4_ratio),
                      by=c("sub"))
all_dat <- inner_join(all_dat, r_dat, by="sub")
head(all_dat)
```

Now I have data, I can do a preliminary analysis.

## Checking how routine was affected by group

The variable group were less routine, as we would expect.

```{r}
all_dat$train_type <- as.factor(all_dat$train_type)
levels(all_dat$train_type) <- c("stable", "variable")
all_dat %>% ggplot(aes(x=log(r), fill=train_type, group=train_type)) + geom_density(alpha=0.5)
```

```{r}
with(all_dat, t.test(log(r) ~ train_type))
```

This seems pretty sensible. Now I want to assess whether the measure of routine predicts the accuracy or k4 difference.

I'll just do a qqplot to see how skewed the measures are.

**First the measure of routine**

```{r}

all_dat %>% ggplot(aes(sample=r)) + stat_qq() + stat_qq_line()
```

A bit hairy. Will try logging to see if that improves things.

```{r}

all_dat %>% ggplot(aes(sample=log(r))) + stat_qq() + stat_qq_line()
```

That improves things a lot. Will keep the r measure logged.

**Now the DVs**

```{r}
all_dat %>% ggplot(aes(sample=acc_ratio)) + stat_qq() + stat_qq_line()
```

Eech! That measure will need some work

```{r}

all_dat %>% ggplot(aes(sample=k4_ratio)) + stat_qq() + stat_qq_line()
```

Okies, its clear we need to log transform the accuracy and k4 measures

```{r}
# first, aremove anyone who got a k4 diff of 0, as this means one of our estimates for them was infinity
all_dat$k4_ratio[!all_dat$k4_ratio] <- NA
all_dat$k4_ratio[is.infinite(all_dat$k4_ratio)] <- NA
all_dat <- all_dat %>% mutate(log_r=log(r),
                              log_acc=log(acc_ratio),
                              log_k4=log(k4_ratio))
head(all_dat)
```

Plot the new variables to make sure they look ok

```{r}
all_dat %>% ggplot(aes(sample=log_acc)) + stat_qq() + stat_qq_line()
```

Euw. Still not amazing, but we can go spearmans if needs be

```{r}

all_dat %>% ggplot(aes(sample=log_k4)) + stat_qq() + stat_qq_line()
```

Ah, much better.

Okies, now remove outliers \> or \< 3 \* sds from the mean

```{r}

all_dat <- all_dat %>% na.omit()
cutoffs <- all_dat %>% summarise(mu_lr = mean(log_r),
                      sd_lr = sd(log_r),
                      mu_la = mean(log_acc),
                      sd_la = sd(log_acc),
                      mu_lk4 = mean(log_k4),
                      sd_lk4 = sd(log_k4))

all_dat <- all_dat %>% mutate(log_r = if_else(log_r < cutoffs$mu_lr + 3*cutoffs$sd_lr &
                                    log_r > cutoffs$mu_lr - 3*cutoffs$sd_lr, log_r, NA),
                   log_acc = if_else(log_acc < cutoffs$mu_la + 3*cutoffs$sd_la &
                                    log_acc > cutoffs$mu_la - 3*cutoffs$sd_la, log_acc, NA),
                   log_k4 = if_else(log_k4 < cutoffs$mu_lk4 + 3*cutoffs$sd_lk4 &
                                    log_k4 > cutoffs$mu_lk4 - 3*cutoffs$sd_lk4, log_k4, NA),)

all_dat <- all_dat %>% na.omit()
```

## Does routine formation predict differences between the learning transfer conditions?

Lets do the pairs plot:

```{r}

ggpairs(all_dat %>% select(train_type, log_r, log_acc, log_k4),
        mapping=ggplot2::aes(colour = all_dat$train_type))
```

Just going to print a scatter plot of the acc ratio / log r relationship for a talk

```{r}

p_wdth = 10
p_hgt = 10

pdf(paste('/home/kgarner/Documents/projects/doors-data/figs/scatter_talk.pdf', sep=""), # for talks
          width = p_wdth/2.54*2.5, height = p_hgt/2.54*2.5)
par(mar = c(5, 5, 2, 1), las=2, cex=2)
with(all_dat, plot(x=log_r, y=log_acc, pch=19,
     frame.plot=F, ylim=c(-0.2, .2),
     xlim=c(0, 4),
     xlab = "log r",
     ylab = "learning",
     col = wes_palette("IsleofDogs1")[1],
     main = "",
     xaxt = "n",
     yaxt = "n",
     cex = 2,
     cex.lab = 2))
axis(1, at = seq(0, 4, by = 2), cex.axis = 1.5)
axis(2, at = seq(-0.2, 2, by = 0.1), cex.axis = 1.5)
with(all_dat, points(x=log_r, y=log_acc, pch=1, cex=2))
# draw regression line
mod <- lm(log_acc ~ log_r, data=all_dat)
abline(mod, col = "black", lwd = 1)
text(x=1, y=-0.15, labels="r(81)=-0.23***", cex = 1.5)
dev.off()

```
